---
title: "Introduction to Regression"
subtitle: "Predicting Campus Chaos"
format:
  revealjs:
    css: style_lecture_purple.css
    html-math-method: mathjax
    slide-number: true
    transition: fade
    background-transition: fade
    highlight-style: pygments
    width: 1200
    height: 800
    margin: 0.1
    min-scale: 0.5
    max-scale: 2.0
execute:
  echo: true
  message: false
  warning: false
  error: true
---


```{r, include=FALSE}
# Setup: Load libraries and fit all models upfront
library(tidyverse)
library(broom)

set.seed(42)

chaos_data <- tibble(
  id = 1:200,
  sleep_hours = rnorm(200, mean = 6.5, sd = 1.5),
  coffee_cups = rnorm(200, mean = 2.8, sd = 1.2),
  tiktok_hours = rnorm(200, mean = 3.2, sd = 2.1),
  distance_campus_miles = rnorm(200, mean = 5, sd = 4),
  group_chats = rnorm(200, mean = 12, sd = 5),
  dining_hall = sample(c("North", "South", "West"), 200, replace = TRUE),
  music_genre = sample(c("K-pop", "Hyperpop"), 200, replace = TRUE),
  major_type = sample(c("STEM", "Non-STEM"), 200, replace = TRUE),
  pet_owner = sample(c("Yes", "No"), 200, replace = TRUE),
  touches_grass = sample(c("Never", "Sometimes", "Regular"), 200, replace = TRUE),
  class_year = sample(c("Fr", "So", "Ju", "Se"), 200, replace = TRUE)
) %>%
  mutate(
    music_hyperpop = ifelse(music_genre == "Hyperpop", 1, 0),
    major_stem = ifelse(major_type == "STEM", 1, 0),
    pet_yes = ifelse(pet_owner == "Yes", 1, 0),
    grass_never = ifelse(touches_grass == "Never", 1, 0),
    grass_regular = ifelse(touches_grass == "Regular", 1, 0),
    class_soph = ifelse(class_year == "Soph", 1, 0),
    class_jun = ifelse(class_year == "Jun", 1, 0),
    class_sen = ifelse(class_year == "Sen", 1, 0),
    chaos_score = 50 + (-2.5 * sleep_hours) + (3.2 * coffee_cups) +
                  (1.8 * tiktok_hours) + (0.8 * distance_campus_miles) +
                  (0.3 * group_chats) + (5 * music_hyperpop) + (-3 * major_stem) +
                  (8 * pet_yes) + (6 * grass_never) +
                  (2 * (dining_hall == "South")) + (-1 * (dining_hall == "West")) +
                  (2 * class_soph) + (4 * class_jun) + (6 * class_sen) +
                  (1.5 * sleep_hours * coffee_cups) + (0.9 * tiktok_hours * pet_yes) +
                  (-0.5 * sleep_hours * major_stem) + rnorm(200, mean = 0, sd = 8)
  )

# Fit all models now
model_sleep <- glm(chaos_score ~ sleep_hours,
                   family = gaussian(link = "identity"), data = chaos_data)

model_sleep_coffee <- glm(chaos_score ~ sleep_hours + coffee_cups,
                          family = gaussian(link = "identity"), data = chaos_data)

model_music <- glm(chaos_score ~ music_hyperpop,
                   family = gaussian(link = "identity"), data = chaos_data)

model_grass <- glm(chaos_score ~ grass_never + grass_regular,
                   family = gaussian(link = "identity"), data = chaos_data)

model_full <- glm(chaos_score ~ sleep_hours + coffee_cups + tiktok_hours +
                  distance_campus_miles + group_chats + music_hyperpop +
                  major_stem + pet_yes + grass_never + grass_regular,
                  family = gaussian(link = "identity"), data = chaos_data)
```

# Regression Fundamentals

## What is Regression?

- We are often interested in how two variables are related.

- **Regression analysis** examines this relationship, where one variable predicts or explains another.

- Simple linear regression:

$$
\hat{y} = \hat{\beta}_0 + \hat{\beta}_1 x_1
$$

- Multiple regression:

$$
\hat{y} = \hat{\beta}_0 + \hat{\beta}_1 x_1 + \hat{\beta}_2 ... + \hat{\beta}_kx_k
$$

## Key Terms

- **Response variable** (outcome variable): The variable we are interested in predicting or modeling

    - This is our $y$ 

- **Explanatory variable** (predictor variable): The variable explaining or predicting the outcome

    - This is our $x$

- In our modeling, we will model a single outcome as a function of predictor(s) variables.

## Campus Chaos Score

- We will model a student's **Campus Chaos Score**: a composite measure of how much delightful chaos they introduce to campus life.

::: {.tab-callout .tab-blue}
::: {.tab-label}
Response Variable
:::

**Campus Chaos Score**: a composite measure of how much delightful chaos a student introduces to campus life (scale: 0-100)
:::

## Campus Chaos Score

- We will model a student's **Campus Chaos Score**: a composite measure of how much delightful chaos they introduce to campus life.

::: {.tab-callout .tab-green}
::: {.tab-label}
Explanatory Variables
:::

**Continuous predictors:**

- Sleep per night (hrs)
- Coffee per day (cups)
- TikTok scrolling per day (hr)
- Distance from campus (miles)
:::

## Campus Chaos Score

- We will model a student's **Campus Chaos Score**: a composite measure of how much delightful chaos they introduce to campus life.

::: {.tab-callout .tab-green}
::: {.tab-label}
Explanatory Variables
:::

**Categorical predictors:**

- Student classification (Fr, So, Ju, Se)
- Major type (STEM vs Non-STEM)
- Music genre (K-pop vs Hyperpop)
- Pet ownership (Yes/No)
:::

# Ordinary Least Squares (OLS)

## The Regression Line

- The **OLS regression line** minimizes the sum of squared distances between observed and predicted values.

$$\hat{y} = \hat{\beta}_0 + \hat{\beta}_1 x$$

- $\hat{\beta}_0$ = intercept
- $\hat{\beta}_1$ = slope
- $\hat{y}$ = predicted value

## The Regression Line



## Simple Regression: Sleep

- Looking at chaos_score ~ sleep_hours,

```{r}
#| echo: false
tidy(model_sleep) |> select(-statistic, -p.value)
```

- The fitted line: 

$$\hat{y} = `r round(coef(model_sleep)[1], 2)` + `r round(coef(model_sleep)[2], 2)` x$$


# Interpreting Coefficients

## The Slope $\hat{\beta}_1$

- Template for interpretations:

    - For a **[k] [unit]** increase in **[x]**, we expect **[y]** to **increase/decrease** by **[k × |b₁|] [units]**.

- From our model: $b_1 = `r round(coef(model_sleep)[2], 2)`$

- **Interpretations:**

    - For a 1-hour increase in sleep, we expect chaos to decrease by `r abs(round(coef(model_sleep)[2], 2))` points.

    - For a 2-hour increase in sleep, we expect chaos to decrease by `r abs(round(2 * coef(model_sleep)[2], 2))` points.

## The Intercept $\hat{\beta}_0$

- Template for interpretations:

    - When **[x] = 0**, we expect the mean of **[y]** to be **[$\hat{\beta}_0$]**.

    - **Caveat:** Only interpret if x = 0 is meaningful!

- From our model: $\hat{\beta}_0 = `r round(coef(model_sleep)[1], 2)`$

- **Interpretation:**

    - When sleep = 0 hours, predicted chaos = `r round(coef(model_sleep)[1], 2)` points.

# Categorical Predictors

## Why Categorical?

- Many variables are categorical (music genre, major type, pet ownership, etc.).

- For $c$ categories, we add $c-1$ dummy variables. This is called reference coding.

- Example: Music genre with 2 categories (K-pop, Hyperpop)

    - Include: Hyperpop = 1 if Hyperpop, 0 otherwise
    - Exclude: K-pop (the **reference group**)

- Coefficients represent differences from the reference group.

## Binary Categorical: Music Genre

```{r}
#| echo: false
tidy(model_music) |> select(-statistic, -p.value)
```

- Intercept = Average chaos for K-pop fans (reference): `r round(coef(model_music)[1], 2)` points

- Hyperpop coefficient: `r round(coef(model_music)[2], 2)` points higher than K-pop fans

## Three-Level Categorical: Grass Touching

```{r}
#| echo: false
tidy(model_grass) |> select(-statistic, -p.value)
```

- **Interpretations:**

    - Intercept = Average for "Sometimes" touchers: `r round(coef(model_grass)[1], 2)` points

    - Never: `r round(coef(model_grass)[2], 2)` points higher than "Sometimes"

    - Regular: `r round(coef(model_grass)[3], 2)` points vs "Sometimes"


## Your Turn

- Plan at least three basic models you want to examine as part of your project.

