---
title: "Regression Analysis<br>Basic Inference"
format:
  revealjs:
    css: style_lecture_purple.css
    html-math-method: mathjax
    slide-number: true
    transition: fade
    background-transition: fade
    highlight-style: pygments
    width: 1200
    height: 800
    margin: 0.1
    min-scale: 0.5
    max-scale: 2.0
execute:
  echo: true
  message: false
  warning: false
  error: true
---

## Introduction

```{r, include=FALSE}
# Setup: Load libraries and fit all models upfront
library(tidyverse)
library(broom)

set.seed(42)

chaos_data <- tibble(
  id = 1:200,
  sleep_hours = pmin(10, pmax(0, round(rnorm(200, mean = 6.5, sd = 1.5)))),
  coffee_cups = pmax(0, round(rnorm(200, mean = 2.8, sd = 1.2))),
  tiktok_hours = pmin(10, pmax(0, round(rnorm(200, mean = 3.2, sd = 2.1)))),
  distance_campus_miles = pmax(0, round(rnorm(200, mean = 5, sd = 4), 1)),
  group_chats = pmin(10, pmax(0, round(rnorm(200, mean = 2, sd = 2)))),
  dining_hall = sample(c("North", "South", "West"), 200, replace = TRUE),
  music_genre = sample(c("K-pop", "Hyperpop"), 200, replace = TRUE),
  major_type = sample(c("STEM", "Non-STEM"), 200, replace = TRUE),
  pet_owner = sample(c("Yes", "No"), 200, replace = TRUE),
  touches_grass = sample(c("Never", "Sometimes", "Regular"), 200, replace = TRUE),
  class_year = sample(c("Fr", "So", "Ju", "Se"), 200, replace = TRUE)
) %>%
  mutate(
    music_hyperpop = ifelse(music_genre == "Hyperpop", 1, 0),
    major_stem = ifelse(major_type == "STEM", 1, 0),
    pet_yes = ifelse(pet_owner == "Yes", 1, 0),
    grass_never = ifelse(touches_grass == "Never", 1, 0),
    grass_regular = ifelse(touches_grass == "Regular", 1, 0),
    class_soph = ifelse(class_year == "Soph", 1, 0),
    class_jun = ifelse(class_year == "Jun", 1, 0),
    class_sen = ifelse(class_year == "Sen", 1, 0),
    chaos_score = 50 + (-2.5 * sleep_hours) + (3.2 * coffee_cups) +
                  (1.8 * tiktok_hours) + (0.8 * distance_campus_miles) +
                  (0.3 * group_chats) + (5 * music_hyperpop) + (-3 * major_stem) +
                  (8 * pet_yes) + (6 * grass_never) +
                  (2 * (dining_hall == "South")) + (-1 * (dining_hall == "West")) +
                  (2 * class_soph) + (4 * class_jun) + (6 * class_sen) +
                  (1.5 * sleep_hours * coffee_cups) + (0.9 * tiktok_hours * pet_yes) +
                  (-0.5 * sleep_hours * major_stem) + rnorm(200, mean = 0, sd = 8)
  )

# Fit all models now
model_full <- glm(chaos_score ~ sleep_hours + coffee_cups + tiktok_hours +
                  distance_campus_miles + group_chats + music_hyperpop +
                  major_stem + pet_yes + grass_never + grass_regular,
                  family = gaussian(link = "identity"), data = chaos_data)
```

- Last week:

    - Regression estimation
    - Interpretations 
    
- Today: 

    - Making inference using regression models
    
- Wednesday: 

    - Other predictor types
    
- Next week: 

    - Leaving the normal distribution

## Model Construction

- Recall our campus chaos dataset,

```{r}
head(chaos_data)
```

## Model Construction

- Today we will work with the basic model

$$\text{chaos} = \beta_0 + \beta_1 \text{ sleep} + \beta_2 \text{ coffee} + \beta_3 \text{ tiktok} + \beta_4 \text{ distance}$$

```{r}
m_full <- glm(chaos_score ~ sleep_hours + 
                            coffee_cups + 
                            tiktok_hours + 
                            distance_campus_miles,
              data = chaos_data)
```

## Model Construction

- Examining results, 

```{r}
coefficients(m_full)
```

$$\text{chaos} = 29.84 + 1.99 \text{ sleep} + 12.76 \text{ coffee} + 2.54 \text{ tiktok} + 0.23 \text{ distance}$$

## Questions we can ask

- Does this regression line significantly predict chaos score?

    - Omnibus test for a significant line

- Do individual predictors predict chaos score?

    - Individual tests for predictors

## Reminder of Statistical Inference

- **hypothesis testing** a statistical procedure that allows us to make inferences about population parameters based on sample data.

- General steps:

    1. Make a statement about the nature of the population. 
    
    2. Collect evidence to test the statement.
    
    3. Analyze the data to determine the plausibility of the statement.
    
## Hypotheses

- **hypothesis** a statement regarding the population parameter.

- We will be stating two hypotheses: the null and the alternative.

    - **$H_0$: null hypothesis** a mathematical statement about the population parameter; this hypothesis is assumed *true* until evidence indicates otherwise.
    
    - **$H_1$: alternative hypothesis** a statement about the population parameter that contradicts the null hypothesis; this is also called the research hypothesis.
    
- Note! $H_0 + H_1$ should cover the real line, $\mathbb{R}$.

    - Some textbooks teach that "$=$" is *always* used in $H_0$.
    
## Test Statistics

- After stating our hypotheses, we will construct an appropriate test statistic. 

    - A test statisic is a measure of the distance from what you observed to what you are assuming to he true.

- The choice of test statistic depends on:

    1. The hypotheses being tested.
    
    2. Assumptions made about the data.

## Test Statistics

- The value of the test statistic depends on the sample data. 

    - **If we were to draw a different sample, our calculations would result in a different value for the test statistic.**

- We use the test statistic on our way to drawing conclusions about the hypotheses.

## *p*-Value

- ***p*-value** the probability of observing our test statistic or one more extreme.

    - More extreme?
        
        - If the alternative hypothesis is right-tailed, we will consider the right tail of the distribution.
        
        - If the alternative hypothesis is left-tailed, we will consider the left tail of the distribution.
        
        - If the alternative hypothesis is two-tailed, we will consider both tails of the distribution.

- We use the *p*-value to determine if we reject the null (believe the alternative) or fail to reject the null (do not believe the alternative).

## *p*-Value

- ***p*-value** the probability of observing our test statistic or one more extreme.

```{=html}
<style>
/* Sizes use the slide's own coordinate system, not the browser viewport. */
#pval-widget, #pval-widget * { box-sizing: border-box; }
#pval-widget {
  width: 100%;
  margin: 0 auto;
  font-size: 22px;
  line-height: 1.25;
}
#pval-widget .pval-layout {
  display: grid;
  grid-template-columns: minmax(0, 30%) minmax(0, 1fr);
  gap: 14px;
  align-items: stretch;
}
#pval-widget .pval-controls,
#pval-widget .pval-plot {
  min-width: 0;
  background: #f5f4f0;
  border-radius: 10px;
  padding: 16px;
}
#pval-widget .pval-heading {
  font-weight: 700;
  margin-bottom: 12px;
}
#pval-widget .pval-option {
  display: flex;
  align-items: center;
  gap: 8px;
  margin-bottom: 10px;
  font-size: inherit;
}
#pval-widget input[type="radio"] {
  margin: 0;
  flex: 0 0 auto;
  accent-color: #005EA6;
}
#pval-widget .pval-statistic {
  display: flex;
  justify-content: space-between;
  gap: 8px;
  margin-top: 18px;
}
#pval-widget #pvalZVal { font-weight: 700; font-variant-numeric: tabular-nums; }
#pval-widget #pvalZ {
  display: block;
  width: 100%;
  min-width: 0;
  margin: 12px 0 0;
  accent-color: #005EA6;
}
#pval-widget .pval-plot {
  display: flex;
  flex-direction: column;
  justify-content: center;
}
#pval-widget #pvalSvg {
  display: block;
  width: 100%;
  height: 220px;
  margin: 0;
}
#pval-widget .pval-result {
  margin-top: 8px;
  text-align: center;
  font-weight: 700;
  font-size: 28px;
  color: #005EA6;
  font-variant-numeric: tabular-nums;
}
</style>
<div id="pval-widget">
  <div class="pval-layout">
    <div class="pval-controls">
      <div class="pval-heading">Alternative hypothesis:</div>
      <label class="pval-option"><input type="radio" name="pvalTail" value="upper"> Upper-tailed</label>
      <label class="pval-option"><input type="radio" name="pvalTail" value="lower"> Lower-tailed</label>
      <label class="pval-option"><input type="radio" name="pvalTail" value="two" checked> Two-tailed</label>
      <div class="pval-statistic">
        <label for="pvalZ">Test statistic</label>
        <span id="pvalZVal">1.50</span>
      </div>
      <input id="pvalZ" type="range" min="-4" max="4" step="0.01" value="1.5">
    </div>
    <div class="pval-plot">
      <svg id="pvalSvg" viewBox="0 0 700 260" preserveAspectRatio="xMidYMid meet" role="img" aria-label="Standard normal distribution with shaded p-value area">
        <line x1="40" y1="220" x2="660" y2="220" stroke="#888" stroke-width="1"/>
        <path id="pvalShadeLeft" fill="#7DB9D9" fill-opacity="0.7" d=""/>
        <path id="pvalShadeRight" fill="#7DB9D9" fill-opacity="0.7" d=""/>
        <path id="pvalCurve" fill="none" stroke="#005EA6" stroke-width="2.5" d=""/>
        <line id="pvalRule" x1="0" y1="20" x2="0" y2="220" stroke="#222" stroke-width="1.5" stroke-dasharray="4,3"/>
      </svg>
      <div class="pval-result" aria-live="polite">p-value = <span id="pvalOut">0.1336</span></div>
    </div>
  </div>
</div>
<script>
(function() {
  function erf(x) {
    const sign = x < 0 ? -1 : 1;
    x = Math.abs(x);
    const a1 = 0.254829592, a2 = -0.284496736, a3 = 1.421413741,
          a4 = -1.453152027, a5 = 1.061405429, p = 0.3275911;
    const t = 1 / (1 + p * x);
    const y = 1 - (((((a5 * t + a4) * t) + a3) * t + a2) * t + a1) * t * Math.exp(-x * x);
    return sign * y;
  }
  function normPdf(x) { return Math.exp(-0.5 * x * x) / Math.sqrt(2 * Math.PI); }
  function normCdf(x) { return 0.5 * (1 + erf(x / Math.sqrt(2))); }
  const xMin = -4, xMax = 4, yMin = 0, yMax = 0.45;
  const pxLeft = 40, pxRight = 660, pyTop = 20, pyBottom = 220;
  function xToPx(x) { return pxLeft + (x - xMin) / (xMax - xMin) * (pxRight - pxLeft); }
  function yToPx(y) { return pyBottom - (y - yMin) / (yMax - yMin) * (pyBottom - pyTop); }
  function curvePoints(x0, x1, step) {
    const pts = [];
    for (let x = x0; x <= x1 + 1e-9; x += step) {
      pts.push([xToPx(x), yToPx(normPdf(x))]);
    }
    return pts;
  }
  function areaPath(x0, x1) {
    const pts = curvePoints(x0, x1, 0.05);
    if (pts.length === 0) return "";
    let d = `M ${pts[0][0]},${yToPx(0)} `;
    pts.forEach(p => { d += `L ${p[0]},${p[1]} `; });
    d += `L ${pts[pts.length - 1][0]},${yToPx(0)} Z`;
    return d;
  }
  const curveEl = document.getElementById('pvalCurve');
  const shadeLeftEl = document.getElementById('pvalShadeLeft');
  const shadeRightEl = document.getElementById('pvalShadeRight');
  const ruleEl = document.getElementById('pvalRule');
  const zSlider = document.getElementById('pvalZ');
  const zValEl = document.getElementById('pvalZVal');
  const pvalOutEl = document.getElementById('pvalOut');
  // Draw the static curve once
  const fullPts = curvePoints(xMin, xMax, 0.05);
  curveEl.setAttribute('d', 'M ' + fullPts.map(p => p.join(',')).join(' L '));
  function getTail() {
    const checked = document.querySelector('input[name="pvalTail"]:checked');
    return checked ? checked.value : 'two';
  }
  function update() {
    const z = parseFloat(zSlider.value);
    zValEl.textContent = z.toFixed(2);
    const tail = getTail();
    let pvalue;
    shadeLeftEl.setAttribute('d', '');
    shadeRightEl.setAttribute('d', '');
    if (tail === 'upper') {
      pvalue = 1 - normCdf(z);
      shadeRightEl.setAttribute('d', areaPath(z, xMax));
    } else if (tail === 'lower') {
      pvalue = normCdf(z);
      shadeLeftEl.setAttribute('d', areaPath(xMin, z));
    } else {
      pvalue = 2 * (1 - normCdf(Math.abs(z)));
      shadeLeftEl.setAttribute('d', areaPath(xMin, -Math.abs(z)));
      shadeRightEl.setAttribute('d', areaPath(Math.abs(z), xMax));
    }
    const zx = xToPx(z);
    ruleEl.setAttribute('x1', zx);
    ruleEl.setAttribute('x2', zx);
    pvalOutEl.textContent = pvalue.toFixed(4);
  }
  zSlider.addEventListener('input', update);
  document.querySelectorAll('input[name="pvalTail"]').forEach(function(r) {
    r.addEventListener('change', update);
  });
  update();
})();
</script>
```

## *p*-Values

- Mathematically, we are asking

    - Right-tailed test: $P[\text{distribution} \ge \text{test statistic}]$
    
    - Left-tailed test: $P[\text{distribution} \le \text{test statistic}]$
    
    - Two-tailed test: $P[\text{distribution} \ne \text{test statistic}]$

## Rejection Regions

- We use rejection regions to determine if we reject the null or fail to reject the null.

- Classical approach: critical values and statistical tables. 

- Modern approach: $p$-values.

    - ***p*-value** the probability of observing our test statistic or one more extreme.

- All rejection regions will take the same form:

<center><span style="font-size:2em;">Reject $H_0$ if $p < \alpha$.</span></center>

## Errors in Hypothesis Testing

- Wait, what is $\alpha$? 

    - $\alpha$ is the **significance level** of the test. 

- A Type I error ($\alpha$) means that we reject the null when we should not. 

    - $\alpha = P[\text{reject } H_0 \text{ when } H_0 \text{ is true}]$

- A Type II error ($\beta$) means that we fail to reject the null when we should not. 

    - $\beta = P[\text{fail to reject } H_0 \text{ when } H_1 \text{ is true}]$
    
- We should choose $\alpha$ based on the level of error we are willing to withstand.

    - The $\alpha$ that is commonly used is $\alpha=0.05$.
    
    - Sometimes, smaller $\alpha$ is used. e.g., clinical trial $\to$  $\alpha=0.01$.

## Drawing a Conclusion 

- We draw a conclusion by following the instructions of our rejection region.

- If $p < \alpha$, we **reject** the null hypothesis.
    - There **is** sufficient evidence to suggest that the alternative is true.

- If $p \ge \alpha$, we **fail to reject** the null hypothesis.
    - There **is not** sufficient evidence to suggest that the alternative is true.

- Note the language -- "fail to reject" is not the same as "accept."    

<br><center><span style="font-size:2em;">We **never** "accept" $H_0$.</span></center>    

## Interpretation of Results

- We will always report our results in the context of the problem.

- Examples:

    - There is sufficient evidence to suggest that the average daily spending of guests at Disney's Hollywood Studios is greater than $50.
    
    - There is not sufficient evidence to suggest that the proportion of guests who purchase Lightning Lane passes differs between annual passholders and single-day ticket holders.

## Putting it Together

- For class purposes, you will be reporting all hypothesis test pieces.

    - Hypotheses
    - Test Statistic and $p$-Value
    - Rejection Region
    - Conclusion and Interpretation

- For poster purposes, we will reduce the amount of information shown about each hypothesis test. 

    - We will draw conclusions and support with a *p*-value.
    
    - e.g., There is sufficient evidence to suggest that the average daily spending of guests at Disney's Hollywood Studios is greater than $50 ($p = 0.012$).

##  Significant Regression Line

- We can use the likelihood ratio test to test for a "significant regression" line. 

    - A significant regression line means that there is a non-zero slope among all slopes... why do we care?

- **Hypotheses**

    - $H_0: \ \beta_1 = ... = \beta_k = 0$ 
    - $H_1:$ at least one is different

- **Test Statistic**

    - $F_0$ 
    
## Significant Regression Line (R)

- We will use the `anova()` function to test for a significant model.

```{r}
#| eval: false
# full model with all predictors
full <- glm(y ~ x_1 + x_2 + ... + x_k, data = dataset_name)
# intercept-only model
reduced <- glm(y ~ 1, data = dataset_name)
# does this particular predictor set explain the outcome?
anova(reduced, full, test = "Chisq")
```

## Significant Regression Line

- In our example model,

```{r}
m_full <- glm(chaos_score ~ sleep_hours + coffee_cups + tiktok_hours + distance_campus_miles, data = chaos_data)
m_reduced <- glm(chaos_score ~ 1, data = chaos_data)
anova(m_reduced, m_full)
```

## Significant Regression Line

- **Hypotheses**

    - $H_0: \beta_1 = \beta_2 = \beta_3 = \beta_4 = 0$ 
    - $H_1:$ at least one is different

- **Test Statistic and *p*-Value**

    - $F_0 = 137.46$
    - $p < 0.001$
    
- **Rejection Region**

    - Reject $H_0$ if $p < \alpha$; $\alpha = 0.05$
    
- **Conclusion and Interpretation **

    - Reject $H_0$. There is sufficient evidence to suggest that at least one of the slopes is non-zero.

##  Individual Predictors

- We can use the $t$-test to test for "significant predictors"

    - A significant predictor means that there is a non-zero slope between the outcome and this specific $x_i$. Why do we care?
    
- **Hypotheses**

    - $H_0: \ \beta_i = 0$ 
    - $H_1: \ \beta_i \ne 0$

- **Test Statistic**

$$t_0 = \frac{\hat{\beta}_i}{\text{SE}_{\beta_i}} \sim t_1$$

## Individual Predictors (`R`)

- We turn to `tidy()` for output,

```{r}
#| eval: false
model_name |> tidy(conf.int = TRUE)
```

- We will see results for *all predictors* in the model. 

## Individual Predictors

- In our example,

```{r}
m_full |> tidy()
```

## Wrap Up

- Now that we know the basics of hypothesis testing in regression, we will expand our knowledge over the next few lectures:

    - Types of predictors
    - Visualization of the model
    - How to handle non-normal outcomes
    
## You Try!

- Fresh data on GitHub (pulled 10 am Mon 10/05)

```{r}
operation_data <- read_csv("https://raw.githubusercontent.com/samanthaseals/IDH4032Fa26/refs/heads/main/analysis_files/data/combined_raw.csv")
```

- Please construct the basic models corresponding to the research questions submitted previously. 

- Begin exploring -- is it a significant model? are there significant predictors? 
